### 2024-07-01 FinalThesis/Chapter_02/Manuscript.tex
Publish PHD

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usetikzlibrary{svg.path}

%------------------------------------------------------------------------------
%  Bibtex vs Biblatex
%------------------------------------------------------------------------------
% While bibtex is very nice to use, it's insanely slow. I therefore work with
% bibtex until the final stage. Set the value below to "true" or "false" to pick
% which one should be used.
\newboolean{usebiblatex}
\setboolean{usebiblatex}{true}

% To remove figures from the document, only keeping their captions
\AtBeginDelayedFloats{%
  \renewcommand{\includegraphics}[2][]{}% Remove \includegraphics functionality
  \renewcommand{\resizebox}[3]{}% Remove \resizebox functionality
}
\setcounter{totalnumber}{10}% Allow up to 10 floats per page
\setcounter{topnumber}{10}% Allow up to 10 floats per page
\setcounter{bottomnumber}{10}% Allow up to 10 floats per page
\renewcommand{\efloatseparator}{\mbox{}}% From the documentation

% Depending on the question, the commands below will prepare the document
% accordingly
\ifthenelse{\boolean{usebiblatex}}{
  %-----------------------------------------------------------------------------
  %  If using biblatex
  %-----------------------------------------------------------------------------
  % Load required packages and refer to correct bibliography
  \usepackage{csquotes}
  \usepackage[
      backend=biber
    , style=apa          % To display Author-year in the text
    , natbib=true        % Necessary to use citep, etc.
    , hyperref=true      % To have clickable links in the references
    , useeditor=false    % Don't add editor to publications
    , sorting=ynt        % Sort by year, name, and title
    , uniquename=false   % Avoid that initials are put
  ]{biblatex}
  \addbibresource{Literature.bib}

  % To include author in link
  \DeclareFieldFormat{citehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{#1}}
  \DeclareFieldFormat{textcitehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{%
    #1%
    \ifbool{cbx:parens}
      {\bibcloseparen\global\boolfalse{cbx:parens}}
      {}}}
  \savebibmacro{cite}
  \savebibmacro{textcite}
  \renewbibmacro*{cite}{%
    \printtext[citehyperref]{%
      \restorebibmacro{cite}%
      \usebibmacro{cite}}}
  \renewbibmacro*{textcite}{%
    \ifboolexpr{
      ( not test {\iffieldundef{prenote}} and
        test {\ifnumequal{\value{citecount}}{1}} )
      or
      ( not test {\iffieldundef{postnote}} and
        test {\ifnumequal{\value{citecount}}{\value{citetotal}}} )
    }
      {\DeclareFieldAlias{textcitehyperref}{noformat}}
      {}%
    \printtext[textcitehyperref]{%
      \restorebibmacro{textcite}%
      \usebibmacro{textcite}}}

  % Remove editor form everything but books
  \AtEveryBibitem{%
   \ifentrytype{book}{}{%
    \clearname{editor}
   }
  }
}{
  %-----------------------------------------------------------------------------
  %  If using bibtex
  %-----------------------------------------------------------------------------
  % Load required packages and select citation style
  \usepackage[round]{natbib}
  \bibliographystyle{apalike}
}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Link colors
% \definecolor{linkcolor}{HTML}{000000}
% \hypersetup{
%   colorlinks = true,
%   linkcolor  = linkcolor,
%   urlcolor   = linkcolor,
%   citecolor  = linkcolor,
% }

% Changing the style of captions in figures etc.
\captionsetup{labelfont=bf, format=plain, font=small}

% Change how equations are referenced
\renewcommand{\theequation}{Equation \arabic{equation}}%

% Avoid skip when including text from external .tex file
\newcommand{\inputy}[1]{\input{#1}\unskip}

% For supplementary material
\newcommand{\beginappendix}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{equation}{0}
  \renewcommand{\theequation}{Equation S\arabic{equation}}%
  \setcounter{section}{0}
  \renewcommand{\thesection}{A.\arabic{section}}%
}

%------------------------------------------------------------------------------
%  ORCID
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

% Create the ORCID logo
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol]
      svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3
        ,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white]
      svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
      svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8
        ,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5
        ,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
      svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6
      ,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}

% Command to create the ORCID
\newcommand\orcid[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
  \pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
% \title{Flooding of the Okavango Delta influences Connectivity for Dispersing
% African Wild Dogs}
\title{Dispersal and Connectivity in Increasingly Extreme Climatic Conditions}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1,2} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

% Reduce spacing between authors
\makeatletter
\def\and{%
  \end{tabular}%
  \hskip -0.5em \@plus.17fil\relax
  \begin{tabular}[t]{c}}
\makeatother

% Current Date
% \date{\today}

% And here the masterpiece begins
\begin{document}

% Change page numbering
\pagenumbering{gobble}

% Create Titlepage
\maketitle

%------------------------------------------------------------------------------
%	Titlepage: Additional Info
%------------------------------------------------------------------------------
\begin{flushleft}

\vspace{0.5cm}

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

\textsuperscript{\S} Corresponding author: \href{mailto://david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

\vspace{4cm}

\textbf{Running Title:} Dispersal in a Changing Climate

\vspace{0.5cm}

\textbf{Keywords:} climate change, conservation, dispersal, flood extremes,
individual-based simulations, landscape connectivity, movement ecology,
step-selection functions

\end{flushleft}

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
While climate change has been shown to impact several life-history traits of
wild-living animal populations, little is known about its effects on dispersal
and connectivity.

@@

Across the entire ecosystem, surface water coverage during maximum flood extent
reduces dispersal success (i.e., the propensity of individuals to disperse
between adjacent subpopulations) by 12\% and increases dispersal durations by
17\%. Locally, however, dispersal success diminishes by as much as 78\%.
Depending on the flood extent, alternative dispersal corridors emerge, some of
which in the immediate vicinity of human-dominated landscapes. Notably, under
maximum flood extent, the number of dispersing trajectories moving into
human-dominated landscapes decreases by 41\% at the Okavango Delta's inflow, but
increases by 126\% at the Delta's distal end. This may drive the amplification
of human-wildlife conflict.

@@

\end{abstract}

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
\newpage

% \onehalfspacing
% \tableofcontents
% \doublespacing

% Change page numbering
\newpage
\pagenumbering{arabic}

% Create linenumbers
\linenumbers

\section{Introduction}

@@

Predicting how dispersal and connectivity respond to environmental change
remains challenging \citep{Littlefield.2019}. This is mainly due to insufficient
data on dispersing animals at the appropriate spatial and temporal scales
\citep{Graves.2014, Vasudev.2015}, coupled with limited insights into potential
differences in environmental conditions under climate change
\citep{Scheiter.2009, IPCC.2022}. To address these shortcomings, one approach
has been to combine climate projections and species distribution models with the
aim of predicting future species ranges and investigating the impacts of such
range shifts on structural connectivity \citep{Wasserman.2012, Ashrafzadeh.2019,
Luo.2021}. However, this approach fails to translate predicted atmospheric
conditions into ground-level landscape characteristics (e.g., vegetation cover,
surface-water). An alternative approach, albeit not primarily focusing on
climate change, is to investigate how seasonality affects functional
connectivity (e.g., \citealp{Mui.2017, Osipova.2019, Zeller.2020, Kaszta.2021}).
This is usually achieved using seasonally updated resistance surfaces that
depict the ease or difficulty at which the focal species can traverse a
particular environment in a specific season \citep{Zeller.2012}. Despite its
biological relevance for understanding seasonal variability, this approach
suffers from the short time-span at which processes are investigated and hinders
drawing inferences on the long-term effects of climate change on dispersal and
connectivity. Recently, alternative approaches for investigating landscape
connectivity that combine empirical GPS movement data with a modeling framework
to reconstruct and predict movement trajectories at the level of the individual
have been suggested \citep{Signer.2017, Hofmann.2023, Signer.2023}. Such
predicted trajectories are uncoupled from any temporal constraint and therefore
suitable to investigate the effects of changing landscapes, as observed or
predicted under the influence of climate change, on dispersal and connectivity.
Given that climate change will shift systems towards conditions that currently
represent extremes \citep{Stott.2016, Ummenhofer.2017, IPCC.2022}, we argue that
a focus on extreme, rather than seasonal, variations could serve as a robust way
to learn about the impacts of climate change on patterns of dispersal and
connectivity. Such an approach appears particularly useful in cases where data
on future environmental conditions are difficult to obtain or plagued by
uncertainty \citep{Collins.2012}.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/GraphicalAbstract.pdf}
  \caption{Conceptual illustration of the employed analytical framework to
  investigate connectivity for dispersing African wild dogs under a minimum and
  maximum flood extent in the Okavango Delta. (a) Based on a series of 700
  floodmaps, obtained between the years 2000 and 2019, we derived extreme
  scenarios for a minimum and maximum flood, using the minimum 100 and maximum
  100 historically observed flood extremes. These layers were then joined with a
  static set of covariates and fed into a dispersal simulation. (b) The
  dispersal simulation was based on a previously parametrized integrated
  step-selection function for dispersing wild dogs and allowed simulating
  dispersal trajectories in the two extreme scenarios. (c) From simulated
  dispersal trajectories we generated four summary maps, highlighting
  connectivity patterns and areas potentially prone to human-wildlife conflict.}
  \label{Concept}
 \end{center}
\end{figure}

@@

To investigate the effects of flood extremes on dispersal and connectivity, we
focused on a \textit{core study area} of 70,000 km\textsuperscript{2} that
comprised the Delta and its surroundings (\Cref{StudyArea}b; purple circle). To
accommodate for the long-distance dispersal events observed in this ecosystem
\citep{Cozzi.2020, Hofmann.2021, Cozzi.2023}, we also considered an
\textit{extended study area} of  300,000 km\textsuperscript{2} spanning from
21\degree 30' S to 17\degree 30' S to and 20\degree 30' E to 26\degree E and
comprising the Delta and parts of Angola, Namibia, Botswana, Zimbabwe, and
Zambia (\Cref{StudyArea}a). Rainfall in this area averages at 450 mm and is
mainly concentrated to the wet season between November and April
\citep{Mendelsohn.2004}. Even though local rains and ground table levels
resulting from floods of previous years may impact annual flood levels
\citep{McCarthy.2006}, the Delta's flood extent is primarily driven by
precipitation patterns in its catchment areas in the Angolan highlands, from
where water is channeled into the Cubango and Cuito rivers and subsequently
discharged into the Okavango Delta \citep{McCarthy.1997, McCarthy.2003}. Since
water only slowly descends from the catchment areas into the Delta and across
its shallow alluvial fan (gradient $\sim$ 1/3,400, \citealp{McCarthy.1997}), the
flood usually reaches its maximum extent long after the local rains have ceased,
during the peak dry season in July or August. Once the floodwater reaches the
Delta's distal ends, it percolates at the Thamalakane and Kunyere Faults, two
natural fault lines at which the water-flow is hindered (\Cref{StudyArea}b).
After reaching its maximum extent, water evaporates or percolates and the flood
steadily retracts over subsequent months. Such inner-annual dynamics are further
amplified or buffered by multi-annual cycles in precipitation patterns across
Angola \citep{Wolski.2012}. Due to the intricate interplay between precipitation
patterns in Angola, local ground-table levels, evapotranspiration, and
anthropogenic water abstractions, predictions of the flood extent under climate
change have proven challenging and range from ``much drier'' to ``much wetter''
depending on employed climate scenarios \citep{Kgathi.2006, Wolski.2008,
Hughes.2011, Wolski.2012, Moses.2018}. In the core study area, the vegetation is
dominated by mopane forest (\textit{Colophospermum mopane}), mixed woodland
acacia-dominated (\textit{Acacia spp.}), and grassland \citep{Mendelsohn.2010}.
Human influence is relatively low and mainly concentrated around small villages
at the western and southern periphery of the Delta. The largest urban center is
Maun, a spread-out settlement at the south-eastern tip of the Delta. Large
portions of the core study area are designated as national parks, game reserves,
or wildlife management areas. Some remaining sections of land are used for
cattle farming and suffer from moderate levels livestock depredation and
associated human-wildlife conflict \citep{Gusset.2009, McNutt.2017}. The core
study area and its extended surroundings are part of the world's largest
transboundary conservation initiative, the Kavango-Zambezi Transfrontier
Conservation Area (KAZA-TFCA, \Cref{StudyArea}a), which attempts to re-establish
connectivity among several core habitats. Previous studies have indicated that
this initiative has high potential for improving connectivity among
subpopulations of various species \citep{Brennan.2020, Lines.2021}, including
African wild dogs \citep{Hofmann.2021}. However, a better understanding of
connectivity within this ecosystem under climate change is lacking.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/StudyArea2.png} \caption{(a)
  Location of the study area, which forms part of the Kavango-Zambezi
  Transfrontier Conservation Area (KAZA-TFCA, red polygon) in Southern Africa.
  (b) The division of the core study area (enclosed by the purple ellipse) into
  sub-areas (1 to 6) was based on the hydrographic structure of the Okavango
  Delta and its tributaries. We simulated dispersal trajectories starting at
  random locations within each of the six source areas (orange circles). Purple
  zones (7 to 14) represent zones that we used to identify if and where
  simulated dispersers left the close surroundings of the Okavango Delta (thus
  called egression zones). These zones were generated using a set of cutlines
  (purple dotted lines) originating from the center of the Delta that dissected
  an elliptical buffer surrounding the Delta into sections of equal size and in
  accordance with cardinal directions. The yellow lines represent GPS
  trajectories of dispersing wild dogs that were observed using GPS collars.
  While not all of these individuals were used to parametrize the dispersal
  model presented by \citet{Hofmann.2023}, they serve to highlight the spatial
  scale at which the species disperses. Map lines delineate study areas and do
  not necessarily depict accepted national boundaries.}
  \label{StudyArea}
 \end{center}
\end{figure}

@@

We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m x
250 m. The set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a composite human influence layer representing
settlements, roads, and agricultural areas. A detailed description of these
layers is provided by \citet{Hofmann.2021}. However, unlike Hofmann and
colleagues, who used a single, static flood map, we employed composite flood
maps to study connectivity. Specifically, we generated water-cover layers using
MODIS Terra MCD43A4 satellite imagery that we classified into binary water-cover
maps using an algorithm developed by \citet{Wolski.2017} and implemented in
\texttt{R} (\url{https://github.com/DavidDHofmann/floodmapr}). This algorithm
allowed us to generate weekly updated ``flood extent maps'', thus providing
detailed information about the flood at any given point in time. In total, we
generated 700 flood extent maps, covering the years 2000 to 2019. We used these
maps to produce minimum and maximum flood scenarios, representative of dry and
wet climatic conditions, assuming that climate change will shift the system
towards one of these extremes in the long run. To create the baseline minimum
(and respectively, maximum) flood scenario, we averaged the 100 flood extent
maps with the smallest (highest) flood extent and generated a binary layer by
masking all pixels that were inundated in less than 50\% of the maps (see also
\Cref{Concept}). By averaging across 100 flood extent maps, we followed a
conservative approach and mitigated chances of misrepresenting minimum and
maximum flood extent due to inaccuracies or artifacts in single remote sensed
flood extent maps. The final minimum and maximum flood extent maps
(\Cref{FloodExtent}) show that the flood in the two scenarios covers 3,500
km\textsuperscript{2} and 9,500 km\textsuperscript{2}, respectively. Finally, we
combined the two flood extent maps with the set of above-mentioned habitat
layers into two spatial stacks, each representative of one scenario
(\Cref{Concept}). To prevent edge effects during the dispersal simulation, we
followed \citet{Koen.2010} and expanded the stacks to an extent that was 20\%
wider and taller than the original, filling the so created buffer zone (gray
buffer in \Cref{StudyArea}b) with randomized values from the respective layers.
This would allow simulated dispersers to leave and re-enter the main study area
via randomized buffer zones.

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/FloodExtent.png}
  \caption{(a) Minimum and (b) maximum flood extent. The two maps were generated
  based on 100 minimum and 100 maximum flood extent maps from a series of 700
  almost weekly updated remote sensed MODIS MCD43A4 satellite images spanning
  the years 2000 to 2019.}
  \label{FloodExtent}
 \end{center}
\end{figure}

@@

To compute spatially mapped betweenness scores, we overlaid the study area with
a grid that had a resolution of 2.5 km x 2.5 km and computed the frequency at
which simulated individuals transitioned from one grid-cell to another. A
cell-transition was said to occur whenever a simulated step crossed from one
grid-cell across or into another. In case the same individual repeatedly
realized the same cell-transition, we only counted a single transition to avoid
overemphasizing regions where individuals moved in circles. This procedure
resulted in a weighted edge-list that we used to compute weighted betweenness
scores for each grid-cell, i.e. the importance of the respective grid-cell in
facilitating movement into adjacent areas \citep{Bastille-Rousseau.2018,
Bastille-Rousseau.2021}. We computed betweenness using the \texttt{igraph}
\texttt{R}-package \citep{Csardi.2006}. Calculations of betweenness scores are
computationally demanding and result in discrete and narrow dispersal corridors
\citep{Hofmann.2023}. A higher map resolution thus results in increased
computation time and narrower corridors. In our case, a grid size of 2.5 km x
2.5 km provided a sensible compromise between computational efficiency and
biological relevance of emerging corridors (see also the appendix of
\citealp{Hofmann.2023}).

@@

\subsection{Impacts of Extreme Environmental Conditions on Dispersal}

@@

\subsection{Methodological Limitations and Future Research Directions}

@@

\subsection{Conclusion and Implications}

>>>>

\documentclass[../FinalThesis.tex]{subfiles}
\begin{document}
\chapter{Dispersal and Connectivity in Increasingly Extreme Climatic Conditions}
\label{ChapterFlood}
\thispagestyle{empty}
\vspace{-1cm}
\noindent\hfil\rule{0.75\textwidth}{.4pt}\hfil

\begin{center}
David D. Hofmann \orcid{0000-0003-3477-4365},
Dominik M. Behr \orcid{0000-0001-7378-8538},
John W. McNutt,
Arpat Ozgul \orcid{0000-0001-7477-2642}, and
Gabriele Cozzi \orcid{0000-0002-1744-1940}

% \begin{flushleft}

% \vspace{5cm}
\vfill

Published in Global Change Biology (2024):
\url{https://doi.org/10.1111/gcb.17299}.

% \textsuperscript{1} Department of Evolutionary Biology and Environmental
% Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
% Switzerland.
%
% \vspace{0.25cm}
%
% \textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
% Private Bag 13, Maun, Botswana.
%
% \vspace{0.25cm}
%
% \textsuperscript{\S} Corresponding author:
% \href{mailto:david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

\end{center}

% \vfill

% \textbf{Running Title:}\\African Wild Dog Dispersal in a Changing Climate:
% Lessons Learned from Seasonal Flood Extremes
%
% \vspace{0.5cm}
%
% \textbf{Keywords:}\\movement, connectivity, climate change, seasonality,
% dispersal, conservation, individual-based simulations, step-selection

% \end{flushleft}

\newpage
\section*{Abstract}
While climate change has been shown to impact several life-history traits of
wild-living animal populations, little is known about its effects on dispersal
and connectivity.

@@

Across the entire ecosystem, surface water coverage during maximum flood extent
reduces dispersal success (i.e., the propensity of individuals to disperse
between adjacent subpopulations) by
\inputy{Figures/NumberReachingOtherSourceAreasMaxFloodPercent.tex}\% and
increases dispersal durations by
\inputy{Figures/DurationReachingOtherSourceAreasMaxFloodPercent.tex}\%. Locally,
however, dispersal success diminishes by as much as
\inputy{Figures/NumberReaching6MaxFloodPercent.tex}\%. Depending on the flood
extent, alternative dispersal corridors emerge, some of which in the immediate
vicinity of human-dominated landscapes. Notably, under maximum flood extent, the
number of dispersing trajectories moving into human-dominated landscapes
decreases by \inputy{Figures/DensityHWCPanhandlePercentage.tex}\% at the
Okavango Delta's inflow, but increases by
\inputy{Figures/DensityHWCMaunPercentage.tex}\% at the Delta's distal end. This
may drive the amplification of human-wildlife conflict.

@@

\newpage
\section{Introduction}

@@

Predicting how dispersal and connectivity respond to environmental change
remains challenging \citep{Littlefield.2019}. This is mainly due to insufficient
data on dispersing animals at the appropriate spatial and temporal scales
\citep{Graves.2014, Vasudev.2015}, coupled with limited insights into potential
differences in environmental conditions under climate change
\citep{Scheiter.2009, IPCC.2022}. To address these shortcomings, one approach
has been to combine climate projections and species distribution models with the
aim of predicting future species ranges and investigating the impacts of such
range shifts on structural connectivity \citep{Wasserman.2012, Ashrafzadeh.2019,
Luo.2021}. However, this approach fails to translate predicted atmospheric
conditions into ground-level landscape characteristics (e.g., vegetation cover,
surface-water). An alternative approach, albeit not primarily focusing on
climate change, is to investigate how seasonality affects functional
connectivity (e.g., \citealp{Mui.2017, Osipova.2019, Zeller.2020, Kaszta.2021}).
This is usually achieved using seasonally updated resistance surfaces that
depict the ease or difficulty at which the focal species can traverse a
particular environment in a specific season \citep{Zeller.2012}. Despite its
biological relevance for understanding seasonal variability, this approach
suffers from the short time-span at which processes are investigated and hinders
drawing inferences on the long-term effects of climate change on dispersal and
connectivity. Recently, alternative approaches for investigating landscape
connectivity that combine empirical GPS movement data with a modeling framework
to reconstruct and predict movement trajectories at the level of the individual
have been suggested \citep{Signer.2017, Hofmann.2023, Signer.2024}. Such
predicted trajectories are uncoupled from any temporal constraint and therefore
suitable to investigate the effects of changing landscapes, as observed or
predicted under the influence of climate change, on dispersal and connectivity.
Given that climate change will shift systems towards conditions that currently
represent extremes \citep{Stott.2016, Ummenhofer.2017, IPCC.2022}, we argue that
a focus on extreme, rather than seasonal, variations could serve as a robust way
to learn about the impacts of climate change on patterns of dispersal and
connectivity. Such an approach appears particularly useful in cases where data
on future environmental conditions are difficult to obtain or plagued by
uncertainty \citep{Collins.2012}.

@@

\begin{figure}[htpb]
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/GraphicalAbstract.pdf}
  \caption{Conceptual illustration of the employed analytical framework to
  investigate connectivity for dispersing African wild dogs under a minimum and
  maximum flood extent in the Okavango Delta. (a) Based on a series of 700
  floodmaps, obtained between the years 2000 and 2019, we derived extreme
  scenarios for a minimum and maximum flood, using the minimum 100 and maximum
  100 historically observed flood extremes. These layers were then joined with a
  static set of covariates and fed into a dispersal simulation. (b) The
  dispersal simulation was based on a previously parametrized integrated
  step-selection function for dispersing wild dogs and allowed simulating
  dispersal trajectories in the two extreme scenarios. (c) From simulated
  dispersal trajectories we generated four summary maps, highlighting
  connectivity patterns and areas potentially prone to human-wildlife conflict.}
  \label{Concept}
 \end{center}
\end{figure}

@@

To investigate the effects of flood extremes on dispersal and connectivity, we
focused on a \textit{core study area} of 70,000 km\textsuperscript{2} that
comprised the Delta and its surroundings (\Cref{StudyAreaCH2}b; purple circle).
To accommodate for the long-distance dispersal events observed in this ecosystem
\citep{Cozzi.2020, Hofmann.2021, Cozzi.2023}, we also considered an
\textit{extended study area} of  300,000 km\textsuperscript{2} spanning from
21\degree 30' S to 17\degree 30' S to and 20\degree 30' E to 26\degree E and
comprising the Delta and parts of Angola, Namibia, Botswana, Zimbabwe, and
Zambia (\Cref{StudyAreaCH2}a). Rainfall in this area averages at 450 mm and is
mainly concentrated to the wet season between November and April
\citep{Mendelsohn.2004}. Even though local rains and ground table levels
resulting from floods of previous years may impact annual flood levels
\citep{McCarthy.2006}, the Delta's flood extent is primarily driven by
precipitation patterns in its catchment areas in the Angolan highlands, from
where water is channeled into the Cubango and Cuito rivers and subsequently
discharged into the Okavango Delta \citep{McCarthy.1997, McCarthy.2003}. Since
water only slowly descends from the catchment areas into the Delta and across
its shallow alluvial fan (gradient $\sim$ 1/3,400, \citealp{McCarthy.1997}), the
flood usually reaches its maximum extent long after the local rains have ceased,
during the peak dry season in July or August. Once the floodwater reaches the
Delta's distal ends, it percolates at the Thamalakane and Kunyere Faults, two
natural fault lines at which the water-flow is hindered (\Cref{StudyAreaCH2}b).
After reaching its maximum extent, water evaporates or percolates and the flood
steadily retracts over subsequent months. Such inner-annual dynamics are further
amplified or buffered by multi-annual cycles in precipitation patterns across
Angola \citep{Wolski.2012}. Due to the intricate interplay between precipitation
patterns in Angola, local ground-table levels, evapotranspiration, and
anthropogenic water abstractions, predictions of the flood extent under climate
change have proven challenging and range from ``much drier'' to ``much wetter''
depending on employed climate scenarios \citep{Kgathi.2006, Wolski.2008,
Hughes.2011, Wolski.2012, Moses.2018}. In the core study area, the vegetation is
dominated by mopane forest (\textit{Colophospermum mopane}), mixed woodland
acacia-dominated (\textit{Acacia spp.}), and grassland \citep{Mendelsohn.2010}.
Human influence is relatively low and mainly concentrated around small villages
at the western and southern periphery of the Delta. The largest urban center is
Maun, a spread-out settlement at the south-eastern tip of the Delta. Large
portions of the core study area are designated as national parks, game reserves,
or wildlife management areas. Some remaining sections of land are used for
cattle farming and suffer from moderate levels livestock depredation and
associated human-wildlife conflict \citep{Gusset.2009, McNutt.2017}. The core
study area and its extended surroundings are part of the world's largest
transboundary conservation initiative, the Kavango-Zambezi Transfrontier
Conservation Area (KAZA-TFCA, \Cref{StudyAreaCH2}a), which attempts to
re-establish connectivity among several core habitats. Previous studies have
indicated that this initiative has high potential for improving connectivity
among subpopulations of various species \citep{Brennan.2020, Lines.2021},
including African wild dogs \citep{Hofmann.2021}. However, a better
understanding of connectivity within this ecosystem under climate change is
lacking.

\begin{figure}[htpb]
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/StudyArea2.png}
  \caption{(a) Location of the study area, which forms part of the
  Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA, red polygon) in
  Southern Africa. (b) The division of the core study area (enclosed by the
  purple ellipse) into sub-areas (1 to 6) was based on the hydrographic
  structure of the Okavango Delta and its tributaries. We simulated dispersal
  trajectories starting at random locations within each of the six source areas
  (orange circles). Purple zones (7 to 14) represent zones that we used to
  identify if and where simulated dispersers left the close surroundings of the
  Okavango Delta (thus called egression zones). These zones were generated using
  a set of cutlines (purple dotted lines) originating from the center of the
  Delta that dissected an elliptical buffer surrounding the Delta into sections
  of equal size and in accordance with cardinal directions. The yellow lines
  represent GPS trajectories of dispersing wild dogs that were observed using
  GPS collars. While not all of these individuals were used to parametrize the
  dispersal model presented by \citet{Hofmann.2023}, they serve to highlight the
  spatial scale at which the species disperses.}
  \label{StudyAreaCH2}
 \end{center}
\end{figure}

@@

We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m x
250 m. The set of layers included \textsf{water-cover, distance-to-water,
tree-cover, shrub/grassland-cover}, and a composite \textsf{human influence}
layer representing settlements, roads, and agricultural areas. A detailed
description of these layers is provided by \citet{Hofmann.2021}. However, unlike
Hofmann and colleagues, who used a single, static flood map, we employed
composite flood maps to study connectivity. Specifically, we generated
water-cover layers using MODIS Terra MCD43A4 satellite imagery that we
classified into binary water-cover maps using an algorithm developed by
\citet{Wolski.2017} and implemented in \texttt{R}
(\url{https://github.com/DavidDHofmann/floodmapr}). This algorithm allowed us to
generate weekly updated ``flood extent maps'', thus providing detailed
information about the flood at any given point in time. In total, we generated
700 flood extent maps, covering the years 2000 to 2019. We used these maps to
produce minimum and maximum flood scenarios, representative of dry and wet
climatic conditions, assuming that climate change will shift the system towards
one of these extremes in the long run. To create the baseline minimum (and
respectively, maximum) flood scenario, we averaged the 100 flood extent maps
with the smallest (highest) flood extent and generated a binary layer by masking
all pixels that were inundated in less than 50\% of the maps (see also
\Cref{Concept}). By averaging across 100 flood extent maps, we followed a
conservative approach and mitigated chances of misrepresenting minimum and
maximum flood extent due to inaccuracies or artifacts in single remote sensed
flood extent maps. The final minimum and maximum flood extent maps
(\Cref{FloodExtent}) show that the flood in the two scenarios covers
\inputy{Figures/FloodExtentMinimum.tex} km\textsuperscript{2} and
\inputy{Figures/FloodExtentMaximum.tex} km\textsuperscript{2}, respectively.
Finally, we combined the two flood extent maps with the set of above-mentioned
habitat layers into two spatial stacks, each representative of one scenario
(\Cref{Concept}). To prevent edge effects during the dispersal simulation, we
followed \citet{Koen.2010} and expanded the stacks to an extent that was 20\%
wider and taller than the original, filling the so created buffer zone (gray
buffer in \Cref{StudyAreaCH2}b) with randomized values from the respective
layers. This would allow simulated dispersers to leave and re-enter the main
study area via randomized buffer zones.

\begin{figure}[htpb]
 \begin{center}
  \includegraphics[width = \textwidth]{Figures/FloodExtent.png}
  \caption{(a) Minimum and (b) maximum flood extent. The two maps were generated
  based on 100 minimum and 100 maximum flood extent maps from a series of 700
  almost weekly updated remote sensed MODIS MCD43A4 satellite images spanning
  the years 2000 to 2019.}
  \label{FloodExtent}
 \end{center}
\end{figure}

@@

To compute spatially mapped betweenness scores, we overlaid the study area with
a grid that had a resolution of 2.5 km x 2.5 km and computed the frequency at
which simulated individuals transitioned from one grid-cell to another. A
cell-transition was said to occur whenever a simulated step crossed from one
grid-cell across or into another. In case the same individual repeatedly
realized the same cell-transition, we only counted a single transition to avoid
overemphasizing regions where individuals moved in circles. This procedure
resulted in a weighted edge-list that we used to compute weighted betweenness
scores for each grid-cell, i.e. the importance of the respective grid-cell in
facilitating movement into adjacent areas \citep{Bastille-Rousseau.2018,
Bastille-Rousseau.2020}. We computed betweenness using the \texttt{igraph}
\texttt{R}-package \citep{Csardi.2006}. Calculations of betweenness scores are
computationally demanding and result in discrete and narrow dispersal corridors
\citep{Hofmann.2023}. A higher map resolution thus results in increased
computation time and narrower corridors. In our case, a grid size of 2.5 km x
2.5 km provided a sensible compromise between computational efficiency and
biological relevance of emerging corridors (see also the appendix of
\citealp{Hofmann.2023}).

@@

% \subsection{Impacts of Extreme Environmental Conditions on Dispersal}

@@

% \subsection{Methodological Limitations and Future Research Directions}

@@

% \subsection{Conclusion and Implications}
